---
output:
  html_document:
    toc: true
    theme: flatly
---

## Introduction

We will compare commands using the [dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) package, and the equivalent generic R commands.

First, let's load the data and the [Magrittr](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) library:

```{r, eval=TRUE, echo=-(1:3)}
library(knitr)
opts_chunk$set(results="hold", cache=TRUE)
options(dplyr.print_min = 3)
url <- "http://cran-logs.rstudio.com/2015/2015-03-14.csv.gz"
file <- paste0(tempdir(), "/../", "temp_file.csv.gz")
if (!file.exists(file)) download.file(url = url, destfile = file, method = "curl")
cran_df <- read.csv(gzfile(file), stringsAsFactors = FALSE)
library(magrittr)
```

Now we load the dplyr library and we create a `tbl_df` variable with the content of `cran_df`: 

```{r, message=FALSE, cache=FALSE}
library(dplyr)
cran_tbl_df <- tbl_df(cran_df)
```

Let's compare both objects and their summaries: 

```{r}
object.size(cran_df); object.size(cran_tbl_df) 
head(cran_df, n = 3); cran_tbl_df
```

## Select

To select all variables from `size` to `country`:

```{r}
# generic R
r <- sapply(c("size", "country"), function(x) which(names(cran_df)==x))
cran_df[,r[1]:r[2]] %>% head(n=3)
# dplyr
cran_tbl_df %>% select(size:country)
# time
system.time(for (i in 1:100) { 
    r <- sapply(c("size", "country"), function(x) which(names(cran_df)==x))
    cran_df[,r[1]:r[2]] } )
system.time(for (i in 1:100) select(cran_tbl_df,size:country))
```

To select all variables excluding those from `time` to `version`:

```{r}
# generic R
r <- sapply(c("time", "version"), function(x) which(names(cran_df)==x))
cran_df[,-(r[1]:r[2])] %>% head(n=3)
# dplyr
cran_tbl_df %>% select(-(time:version))
# time
system.time(for (i in 1:100) {
    r <- sapply(c("time", "version"), function(x) which(names(cran_df)==x))
    cran_df[,-(r[1]:r[2])] } )
system.time(for (i in 1:100) select(cran_tbl_df,-(time:version)))
```

## Filter

To filter all rows for which the package variable is equal to "swirl" or "dplyr": 

```{r}
# generic R
cran_df[ cran_df$package %in% c("swirl","dplyr"), ] %>% head(n=3)
# dplyr
cran_tbl_df %>% filter(package %in% c("swirl","dplyr"))
# time
system.time(for (i in 1:20) cran_df[ cran_df$package %in% c("swirl","dplyr"), ])
system.time(for (i in 1:20) cran_tbl_df %>% filter(package %in% c("swirl","dplyr")))
```

A more complex filter, combined with a select:

```{r}
# generic R
r <- sapply(c("r_version", "version", "r_os"), function(x) which(names(cran_df)==x))
cran_df[ cran_df$r_version >= "3.1" & cran_df$country != "US",
         setdiff(r[1]:r[2],r[3]) ] %>% head(n=3)
# dplyr
cran_tbl_df %>% 
    filter(r_version >= "3.1" & country != "US") %>%
    select(r_version:version, -r_os)
# time
system.time(for (i in 1:10) {
    r <- sapply(c("r_version", "version", "r_os"), function(x) which(names(cran_df)==x))
    cran_df[ cran_df$r_version >= "3.1" & cran_df$country != "US",
             setdiff(r[1]:r[2],r[3]) ] %>% head(n=3) } )
system.time(for (i in 1:10) {
    cran_tbl_df %>% filter(r_version >= "3.1" & country != "US") %>%
                    select(r_version:version, -r_os) } )
```

Not only is dplyr's syntax much more compact and readable, it is also significantly faster !

A last example, with a function applied to one of the variables: 

```{r}
# generic R
r <- sapply(c("date", "size"), function(x) which(names(cran_df)==x))
cran_df[ !is.na(cran_df$r_arch) & cran_df$package == "dplyr", 
         -(r[1]:r[2]) ] %>% head(n=3)
# dplyr
cran_tbl_df %>%
    filter(!is.na(r_arch), package=="dplyr") %>%
    select(-(date:size))
# time
system.time(for (i in 1:10) {
    r <- sapply(c("date", "size"), function(x) which(names(cran_df)==x))
    cran_df[ !is.na(cran_df$r_arch) & cran_df$package == "dplyr", 
             -(r[1]:r[2]) ] %>% head(n=3) } )
system.time(for (i in 1:10) {
    cran_tbl_df %>% filter(!is.na(r_arch), package=="dplyr") %>%
                    select(-(date:size)) } )
```

## Arrange

To arrange rows of a subset of cran in descending order of the variable `size` modulo 10 and then in ascending order of the variable `ip_id`: 

```{r}
# generic R
cran2_df <- cran_df[,3:10]
cran2_df[ with(cran2_df, order( -(size%%10) , ip_id) ) , ] %>% head(n=3)
# dplyr
cran2_tbl_df <- select(cran_tbl_df, size:ip_id)
cran2_tbl_df %>% arrange( desc(size%%10), ip_id )
# time
system.time(for (i in 1:5) cran2_df[ with(cran2_df, order( -(size%%10) , ip_id) ) , ] %>% head(n=3))
system.time(for (i in 1:5) cran2_tbl_df %>% arrange( desc(size%%10), ip_id ))
```

## Mutate

To take a subset of columns and a sample of rows, then add 2 new columns with time stamps converted to actual `POSIXct` values and their numeric values:

```{r}
# generic R
r <- sapply(c("r_version", "r_os"), function(x) which(names(cran_df)==x))
cran3_df <- 
    cran_df %>% (function(x) x[sample(nrow(x),size=10000),-(r[1]:r[2])]) %>%
                within(time_2 <- as.POSIXct(time,format="%H:%M:%S")) %>%
                within(time_s <- as.numeric(time_2))
# dplyr
cran3_tbl_df <- 
    cran_tbl_df %>% select(-(r_version:r_os)) %>%
                    sample_n(10000) %>%
                    mutate(time_2 = as.POSIXct(time,format="%H:%M:%S"),
                           time_3 = as.numeric(time_2))
# time
system.time(for (i in 1:10) { cran_df %>% (function(x) x[sample(nrow(x),size=10000),-(r[1]:r[2])]) %>%
                                          within(time_2 <- as.POSIXct(time,format="%H:%M:%S")) %>%
                                          within(time_3 <- as.numeric(time_2)) } )
system.time(for (i in 1:10) { cran_tbl_df %>% select(-(r_version:r_os)) %>%
                                              sample_n(10000) %>%
                                              mutate(time_2 = as.POSIXct(time,format="%H:%M:%S"),
                                                     time_3 = as.numeric(time_2)) } )
```

## Summarize



---

Qgetuality sources on dplyr:

* The canonical source for dplyr, Hadley Wickham's [dplyr repo](https://github.com/hadley/dplyr)
* Kevin Markham's [dplyr tutorials](https://github.com/justmarkham/dplyr-tutorial) on Github

---

R version:
```{r, echo=FALSE}
version[c("platform", "os", "version.string", "nickname")]
```

*Author: Alexandre Halm*  
*Edited with [RStudio](http://www.rstudio.com/products/RStudio/#Desk)*  

These notes were taken from the [swirl programming assignment](https://class.coursera.org/getdata-013/assignment) from [Coursera](http://www.coursera.org)'s [Getting and Cleaning Data](https://class.coursera.org/getdata-013) course.
