---
output:
  html_document:
    toc: true
    theme: flatly
---

### Introduction

We will compare commands using the [dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) package, and the equivalent generic R commands.

First, let's load the data and the [Magrittr](http://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) library:

```{r, eval=TRUE, echo=TRUE}
url <- "http://cran-logs.rstudio.com/2015/2015-03-14.csv.gz"
file <- paste0(tempdir(), "/../", "temp_file.csv.gz")
if (!file.exists(file)) download.file(url = url, destfile = file, method = "curl")
cran_df <- read.csv(gzfile(file), stringsAsFactors = FALSE)
library(magrittr)
```

Now we load the dplyr library, set some cosmetic options, and we create a `tbl_df` variable with the content of `cran_df`: 

```{r, message=FALSE}
library(dplyr)
options(dplyr.print_min = 3)
cran_tbl_df <- tbl_df(cran_df)
```

Let's compare both objects and their summaries: 

```{r}
object.size(cran_df); object.size(cran_tbl_df) 
head(cran_df, n = 3); cran_tbl_df
```

### Select

To select all variables from `size` to `country`:

```{r}
# Generic R
r <- sapply(c("size", "country"), function(x) which(names(cran_df)==x))
cran_df[,r[1]:r[2]] %>% head(n=3)
# dplyr
cran_tbl_df %>% select(size:country)
# time
system.time(for (i in 1:100) { 
    r <- sapply(c("size", "country"), function(x) which(names(cran_df)==x))
    x <- cran_df[,r[1]:r[2]] } )
system.time(for (i in 1:100) x <- select(cran_tbl_df,size:country))
```

To select all variables excluding those from `time` to `version`:

```{r}
# Generic R
r <- sapply(c("time", "version"), function(x) which(names(cran_df)==x))
cran_df[,-(r[1]:r[2])] %>% head(n=3)
# dplyr
cran_tbl_df %>% select(-(time:version))
# time
system.time(for (i in 1:100) {
    r <- sapply(c("time", "version"), function(x) which(names(cran_df)==x))
    x <- cran_df[,-(r[1]:r[2])] } )
system.time(for (i in 1:100) x <- select(cran_tbl_df,-(time:version)))
```

### Filter

To filter all rows for which the package variable is equal to "swirl": 

```{r}
# Generic R
cran_df[ cran_df$package == "swirl", ] %>% head(n=3)
# dplyr
cran_tbl_df %>% filter(package == "swirl")
# time
system.time(for (i in 1:20) x <- cran_df[ cran_df$package == "swirl", ])
system.time(for (i in 1:20) x <- filter(cran_tbl_df, package == "swirl"))
```

A more complex filter, combined with a select:

```{r}
# Generic R
r <- sapply(c("r_version", "version", "r_os"), function(x) which(names(cran_df)==x))
cran_df[ cran_df$r_version >= "3.1" & cran_df$country != "US",
         setdiff(r[1]:r[2],r[3]) ] %>% head(n=3)
# dplyr
cran_tbl_df %>% 
    filter(r_version >= "3.1" & country != "US") %>%
    select(r_version:version, -r_os)
#time
system.time(for (i in 1:20) {
    x <- cran_df[ cran_df$r_version >= "3.1" & cran_df$country != "US",
                  setdiff(r[1]:r[2],r[3]) ] %>% head(n=3) } )
system.time(for (i in 1:20) {
    x <- cran_tbl_df %>% filter(r_version >= "3.1" & country != "US") %>%
                         select(r_version:version, -r_os) } )
```

Not only is dplyr's syntax much more readble, it is also significantly faster !



---
R version:
```{r, echo=FALSE}
version
```

*Author: Alexandre Halm*  
*Edited with [RStudio](http://www.rstudio.com/products/RStudio/#Desk)*  

These notes were taken from the [swirl programming assignment](https://class.coursera.org/getdata-013/assignment) from [Coursera](http://www.coursera.org)'s [Getting and Cleaning Data](https://class.coursera.org/getdata-013) course.
