---
output:
  html_document:
    toc: true
    theme: flatly
---    

# R commands and examples


## R environment

### Moving around

To set R's working directory: `setwd("directory_name")`.

### Inspecting objects

`str` displays a compact description of any object, while `summary` displays a more analystic summary of data objects: 

```{r, echo=-(1:2)}
library(knitr)
opts_chunk$set(results="hold", cache=TRUE)
str(c(T,F,F,F,T)); str(str)
x <- runif(33,0,2)
str(x); summary(x)
```

`?function` opens the man page for function.

### Reading and writing data

The canonical function is `read.table`:

```{r}
args(read.table)
```

`read.csv` and `read.csv2` are calls to `read.table` with different default values for the `header`, `sep`, `dec`, `fill` and `comment.char` variables).

The help page for read.table `?read.table` has useful hints to load large tables efficiently.

To write data into a text file while *preserving metadata*, use `dump`/`source` and `dput`/`dget`:

```{r, eval=FALSE}
x <- data.frame(X=1:4, Y=6:9); y <- rnorm(n = 10, mean = 1, sd = 0.1)
dump(c("x","y"), file="newData.R")  # to store x and y in the newData.R file
source("newData.R")  # to load the content of newData.R into R
```

Data is read through *connections*. To open a connection to a file: `file("filename")`, to a url: `url("address")`.


## Manipulating data

### Object names

Giving names to elements of a vector:

```{r}
x <- 1:3
names(x) <- c("foo", "bar", "baz")
x; names(x)
```

For a list: 

```{r}
x <- list(a = 1, b = "Bob")
x; names(x); x$a
```

For a matrix: 

```{r}
m <- matrix(1:4, nrow=2, ncol=2)
dimnames(m) <- list(c("a", "b"), c(123,456))
m
```

### Data frames

Creating a data frame: 

```{r, eval=TRUE, echo=TRUE}
x <- data.frame(foo = 1:3, bar = c(T,F,T), baz = c("Alex", "Bob", "Cathy"))
x; nrow(x); row.names(x); ncol(x); names(x)
```

### Subsetting

`a[...]` always returns an object of the same class as `a` (so list -> list, vector -> vector ...) and can contain one or more elements of `a`.

`a[[...]]` is used to extract *one* element of either a *list* or a *data frame*. It is equivalent to `a$...`.

Some examples: 

```{r}
x <- c("a", "b", "a", "c")
x[2:3]  # numeric index
x=="a"; x[x=="a"]; (1:4)[x=="a"]  # logical index
```

With a list:

```{r}
x <- list(foo = 1:4, bar = c(T,F), baz = "Hello")
x[1:2]
x$bar  # equivalent to x["bar"]
x[c(1,3)]
newVar <- "baz"; x[[newVar]]  # x$newVar would not work
```

With a matrix:

```{r}
m <- matrix(rnorm(6,mean=0,sd=1), 2, 3)
m[1,]; class(m[1,])  # first row (a vector)
m[,2:3]; class(m[,2:3])  # columns 2 to 3
# drop=FALSE forces the output to be a matrix:
class(m[2,3]); class(m[2,3, drop=FALSE])
```

To filter out NA values, use `is.na` and `complete.cases`.x

### Simulation 

Common random variable distributions are accessible through commands called `d*` (density function), `p*` (cumulative distribution), `q*` (quantiles) and `r*` (random numbers). Examples for `*` include `norm`  (normal distribution), `binom` (binomial), `pois` (Poisson), `exp` (exponential), `beta` (Beta), `rchisq` (chi squared) etc.

```{r}
dnorm(x = c(-1.96,0,1.96), mean = 0, sd = 1)  # density
pnorm(q = c(-1.96,0,1.96), mean = 0, sd = 1, lower.tail = TRUE)  # cumulative distribution
qnorm(p = c(0.025,0.5,0.975), mean = 0, sd = 1, lower.tail = TRUE)  # quantiles
rnorm(n = 5)  # generates n random numbers with default parameters
```

To ensure simulation reproducibility, always set the random number seed with `set.seed`: 

```{r}
set.seed(1); rnorm(5); set.seed(1); rnorm(5)
```

Use `sample` to sample from an arbitrary set:

```{r}
sample(letters, size = 5, replace = TRUE)
```


## R programming

### Flow control

`if` returns a value in R:

```{r, eval=FALSE}
y <- if(x > 3) { 10 } else { -10 }  
# but if can also be used in the usual/procedural way:
if(x > 3) { y <- 10 } else { y <- -10 }  
```

`for` syntax: 

```{r}
x <- c(T,F,T,T)
# the following commands are equivalent: 
for(i in 1:4) { print(x[i]) }; for(element in x) { print(element) }
```

### Functions

The following function takes a vector "set" and returns the subset of "set" whose elements are larger than "threshold":

```{r}
subset_with_threshold <- function(set, threshold = 0) {
  set[set>threshold] 
}
subset_with_threshold(-2:2)
```

The function `neg_binom_log_lhd` returns a function, equal to the negative log likelihood, from a sample vector; it can then be passed to the `optimize` function:

```{r}
neg_binom_log_lhd <- function(sample) {
  N <- length(sample)
  n <- sum(sample)
  function(r) { - r^n * (1-r)^(N-n) }
}
sample_1 <- rbinom(10,1,0.8)
negLLL_1 <- neg_binom_log_lhd(sample_1)
optimize(negLLL_1,c(0,1))$minimum
```

### Looping functions

`lapply` applies a function to each element of a list (or of an object that can be cast to a list) and returns a list. `sapply` tries to simplify the output of `lapply` whenever possible:

```{r}
lapply( list(g=rnorm(100,0,1),b=rbinom(100,1,0.5),p=rpois(100,1)) , mean )
sapply( list(g=rnorm(100,0,1),b=rbinom(100,1,0.5),p=rpois(100,1)) , mean )
```

`apply` applies a function to a given margin of a matrix:

```{r}
x <- matrix(1:6, 2, 3); x
apply(x,1,sum)  # the function rowSums does it much faster
apply(x,2,mean)  # the function colMeans does it much faster
y <- matrix( runif(n=999,min=0,max=1) , nrow = 3)
apply(y,1,quantile, probs=c(0.25,0.5,0.75))  # the probs arg is passed to quantile
```

`mapply` allows to call a single function with several sets of arguments: 

```{r}
mapply(rnorm, n=2, sd=runif(3), mean=10*(-1:1))
```

`tapply` applies a function over subsets of a vector:

```{r}
x <- c( rnorm(20,0,1) , rbinom(20,1,0.33) , rpois(20,1) )
f <- gl(n = 3, k = 20, labels = c("norm","binom","pois"))  # a sequence of factors
tapply(x, INDEX=f, FUN=mean)

```

`split` splits an object (e.g. a data frame) along a factor:

```{r}
s <- split(airquality, airquality$Month)
sapply(s, function(x) {colMeans(x[,1:4], na.rm = TRUE)})
```

### Debugging

`traceback()`: traces back the stack of function when called right after an error has occured

`debug(function)`: flags a function for debugging; debugger starts at line 1 at every call of the function

`browser`: can be inserted in a function code; suspends execution of code at that point and enters debugging mode

`options(error = recover)` sets an environment-wide variable; after an error, allows to browse the function call stack

### Profiling

`system.time(exp)` measures the time taken to evaluate exp:

```{r}
system.time(sd(rnorm(10^6)))
```

`Rprof()` keeps track of the function stack at regular time intervals and measures time spent in each function; `summaryRprof()` summarizes the output of `Rprof()`.


## Misc commands and examples

#### Sequences, ranges and repetitions:

```{r}
rep(c(1,2,3),times=3); rep(c(F,T),each=4)
s1 <- seq(from = pi, to = 10, length.out = 5); s1; seq(to = 0, along.with = s1)
```


## R Markdown

First the R Markdown package needs to be installed: 

```{r, eval=FALSE}
install.packages("rmarkdown")
```

To embed an R code chunk, place it between backticks like this: 

```{r, echo=FALSE, comment=NA}
cat("```{r}\n# insert code here\n```")
```

To embed a code chunk *with* code evaluation, specify `{r, eval=TRUE}` (default value,) and to *include* the code:`{r, echo=TRUE}` (default value). For instance, the following code chunk:

```{r, echo=FALSE, comment=NA}
cat("```{R, eval=TRUE, echo=FALSE}\nsummary(cars)\n```")
```

produces: 

```{r, eval=TRUE, echo=FALSE}
summary(cars)
```

---
R version:
```{r, echo=FALSE}
version[c("platform", "os", "version.string", "nickname")]
```

*Author: Alexandre Halm*  
*Edited with [RStudio](http://www.rstudio.com/products/RStudio/#Desk)*  
Notes taken from the [R Programming](https://www.coursera.org/course/rprog) course.
